|

### 2020-10-05 manuscript.Rmd
pushing a first final version

abstract: "@moffatt_inner_2020 reported the results of an experiment (N = 26 in the final sample) comparing the facial (surface) electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference in the perioral muscular activity between the rumination and distraction conditions, @moffatt_inner_2020 concluded that *self-reported* inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. We suggest this conclusion is at best hasty. Indeed, concluding on the absence of an effect based on an under-powered non-significant *p*-value is strongly problematic/uninformative. Moreover, the relation between self-reports and physiological measures was not directly assessed, but only indirectly inferred from differences (or absence thereof) in group averages. Given the ample inter-individual variability in these measures (as suggested by our reanalysis), we think inferring the individual-level relation between self-reports and physiological measures from group averages is inappropriate. Given these limitations, we conclude that there is limited evidence for the main conclusion put forward by @moffatt_inner_2020 and we suggest ways forward, both from a theoretical and from a methodological perspective. Complete source code, reproducible analyses, and figures are available at https://github.com/lnalborczyk/inner_experience_EMG."

@@

To model EMG peak amplitude variations in response to the rumination and distraction inductions, we fitted a Bayesian multivariate regression model with the natural logarithm of the EMG peak amplitude as an outcome and *Condition* (baseline, rumination, distraction) as a categorical predictor. Therefore, the intercept represents the estimated logarithm of the EMG peak amplitude in the baseline condition, and the slopes for the rumination and distraction conditions represent deviations from the baseline. These analyses were conducted using the `brms` package [@R-brms], an `R` implementation of Bayesian multilevel models that employs the probabilistic programming language `Stan` [@carpenter_stan_2017]. We ran four chains including each 10.000 iterations and a warmup of 2.000 iterations. Posterior convergence was assessed examining autocorrelation and trace plots, as well as the Gelman-Rubin statistic. Constant effects estimates were summarised via their posterior mean and 95% credible interval. We also report Bayes factors (BFs) computed using the Savage-Dickey method.^[This method simply consists in taking the ratio of the posterior density at the point of interest divided by the prior density at that point [@wagenmakers_bayesian_2010].] These BFs can be interpreted as updating factors, from prior knowledge (what we knew before seeing the data) to posterior knowledge (what we know after seeing the data). A summary of the estimations from this model is presented in Table \@ref(tab:summary). This analysis revealed strong evidence for the hypothesis of a higher average EMG peak amplitude in the rumination condition as compared to the baseline condition for both the FRO and OOI muscles. However, the BFs supported the null hypothesis (i.e., no difference) between the baseline and distraction conditions for the FRO and were inconclusive for both the OOI and OOS muscles.

@@

## Concluding on the null from under-powered studies: what could go wrong?

@@

" Science has found no proof of intelligent life nearby us in space, therefore intelligent life does not exist nearby us in space."

@@

We suggest the (a priori) power of the study ran by @moffatt_inner_2020 was much lower than suggested by the authors. Indeed, we speculate that the standardised mean difference in EMG peak amplitude between the rumination and distraction conditions may be much weaker than the standardised mean difference in EMG amplitude between the rumination and rest conditions. If we assume that the former is half the size of the latter [which seems reasonable given the high inter-individual variability in such effects, cf. the next section but also @nalborczyk_can_2020], therefore the a priori power of the main statistical test from @moffatt_inner_2020 was around $0.44$, meaning that they had less than 1 chance out of 2 to find a significant effect (given that the effect in the population was actually $0.36$).

@@

While we appreciate the effort, the current approach poses new problems. First, contrary to what the authors suggest, whereas computing a BF indeed allows assessing the *relative evidence* for the null, computing a BF (i.e., comparing two models) does not solve at all the problem of low power. More precisely, the sensitivity (i.e., the ability to attain a certain goal) of an experimental design to detect a given effect is an issue both for frequentist and Bayesian statistical tests. To illustrate this, we present below the results of a simulation aiming to assess the relation between the sample size and the value of the BF.

@@

As shown in Figure \@ref(fig:simulated-power), the natural logarithm of the BF in favour of the alternative hypothesis is growing proportionally with sample size. More precisely, whereas BFs computed on small samples (i.e., below 80 participants) support the null hypothesis, BFs computed on adequately-powered samples support the alternative hypothesis for all three facial muscles. For instance, the average $\text{BF}_{10}$ computed for the OOI muscle with a sample size of 160 participants is of $\exp(2.18) \approx 8.85$, indicating that these data are approximately $8.85$ times more likely under the alternative hypothesis than under the null hypothesis. To sum up, this reveals that although at low sample sizes, the BF may provide (weak) evidence for the null hypothesis (relative to the alternative hypothesis), this pattern may very well reverse for higher sample sizes.

@@

```{r order, fig.cap = "Average log-EMG amplitude by muscle and condition. The black dots and intervals represent the by-group average and 95\\% confidence interval (N = 26). The horizontal white line in the violin plot represents the median. The grey dots represent the individual-level average natural logarithm of the EMG amplitude by muscle and condition."}
# position for groups
pd <- position_dodge(0.8)

@@

Unfortunately, the problems we discussed in the next section about the interpretation of under-powered non-significant results also apply to this test. Namely, obtaining a non-significant effect of group is very weak evidence that order did not play a role in the results, given the (possibly) very low power of the tests that were performed. This statistical argument is supported by the brief visual exploration fo the data that we presented in Figure \@ref(fig:order), which suggests possibly crucial differences between the two groups of participants. However, given the sample size in each group (N = 12 and N = 14), it is impossible to know for sure at this point.

@@

With this paper we aim to nuance the strong conclusion made by @moffatt_inner_2020, who asserted that the inner experience of rumination was not related to its peripheral muscular correlates. First, we reanalysed the data from @moffatt_inner_2020 and provided some nuance to the conclusion that can be made from these data. Second, we discussed the statistical and epistemological reasons that cast doubt upon the main conclusion of @moffatt_inner_2020. Because the tests conducted by @moffatt_inner_2020 were heavily under-powered, they provide only weak evidence for the absence of difference. Third, we highlighted that the order of the conditions participants went through may impact the effects of the rumination induction (although we can not decide on this issue with the present data). Finally, we showed that the group analyses masked important inter-individual variability that should be more carefully examined.

@@

In @nalborczyk_dissociating_2020, we suggested that this observation was coherent with the mental-habit view of depressive rumination [@watkins_habit-goal_2014], which defines rumination as a habitual behaviour, automatically triggered by contextual cues such as negative mood. We know habitual behaviours are more automatic (they are not intentionally initiated) than non-habitual behaviours. It has been observed that the automaticity with which a verbal thought is evoked may influence the degree to which it is enacted, that is, the [e.g., @cohen_motor_1986; @sokolov_inner_1972]...

In the discussion of @nalborczyk_understanding_2019, we discussed these findings in length and proposed several theoretical interpretations that can account for these results...

@nalborczyk_understanding_2019...

@moffatt_inner_2020 stated that:

> "Given that lip muscle activity has previously been related to inner speech production [23], this suggests that rumination has no specific boost of inner speech production relative to distraction. The increase in lip-muscle activity during states where thoughts were guided (rumination, distraction), compared to rest may therefore reflect the extra effort required to guide one’s thoughts and to engage in inner speech production"...

However...

>>>>

abstract: "@moffatt_inner_2020 reported the results of an experiment (N = 26 in the final sample) comparing the facial electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference in the perioral muscular activity between the rumination and distraction conditions, @moffatt_inner_2020 concluded that *self-reported* inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. We suggest this conclusion is at best hasty. Indeed, concluding on the absence of an effect based on an under-powered non-significant *p*-value is strongly problematic/uninformative. Moreover, the relation between self-reports and physiological measures was not directly assessed, but only indirectly inferred from differences (or absence thereof) in group averages. Given the ample inter-individual variability in these measures (as suggested by our reanalysis), we think inferring the individual-level relation between self-reports and physiological measures from group averages is inappropriate. Given these limitations, we conclude that there is limited evidence for the main conclusion put forward by @moffatt_inner_2020 and we suggest ways forward, both from a theoretical and from a methodological perspective. Complete source code, reproducible analyses, and figures are available at https://github.com/lnalborczyk/inner_experience_EMG."

@@

To model EMG peak amplitude variations in response to the rumination and distraction inductions, we fitted a Bayesian multivariate regression model with the natural logarithm of the EMG peak amplitude as an outcome and *Condition* (baseline, rumination, distraction) as a categorical predictor. Therefore, the intercept represents the estimated natural logarithm of the EMG peak amplitude in the baseline condition, and the slopes for the rumination and distraction conditions represent deviations from the baseline. These analyses were conducted using the `brms` package [@R-brms], an `R` implementation of Bayesian multilevel models that employs the probabilistic programming language `Stan` [@carpenter_stan_2017]. We ran four chains including each 10.000 iterations and a warmup of 2.000 iterations. Posterior convergence was assessed examining autocorrelation and trace plots, as well as the Gelman-Rubin statistic. Constant effects estimates were summarised via their posterior mean and 95% credible interval. We also report Bayes factors (BFs) computed using the Savage-Dickey method.^[This method consists in taking the ratio of the posterior density at the point of interest divided by the prior density at that point [@wagenmakers_bayesian_2010].] These BFs can be interpreted as updating factors, from prior knowledge (what we knew before seeing the data) to posterior knowledge (what we know after seeing the data). A summary of the estimations from this model is presented in Table \@ref(tab:summary). This analysis revealed strong evidence for the hypothesis of a higher average EMG peak amplitude in the rumination condition as compared to the baseline condition for both the FRO and OOI muscles. However, the BFs supported the null hypothesis (i.e., no difference) between the baseline and distraction conditions for the FRO and were inconclusive for both the OOI and OOS muscles.

@@

## Concluding on the null hypothesis from under-powered null-hypothesis significance tests: what could go wrong?

@@

@@

We suggest the (a priori) power of the study ran by @moffatt_inner_2020 was much lower than suggested by the authors. Indeed, we speculate that the standardised mean difference in EMG peak amplitude between the rumination and distraction conditions may be much weaker than the standardised mean difference in EMG amplitude between the rumination and rest conditions. If we assume that the former is half the size of the latter [which seems reasonable given the high inter-individual variability in such effects, cf. the next section but also @nalborczyk_can_2020], therefore the a priori power of the main statistical test from @moffatt_inner_2020 was around $0.44$, meaning that they had less than 1 chance out of 2 to find a significant effect (given that the effect in the population was actually $0.36$). In other words, the chances of detecting the effect (was the effect actually non-null) were less than obtaining a head in a coin flip.

@@

While we appreciate the effort, the current approach poses new problems. First, contrary to what the authors suggest, whereas computing a BF indeed allows assessing the *relative evidence* for the null, computing a BF (i.e., comparing two models) does not solve at all the problem of low power. More precisely, the sensitivity (i.e., the ability to attain a certain goal) of an experimental design to detect a given effect is an issue for both frequentist and Bayesian statistical tests. To illustrate this, we present below the results of a simulation aiming to assess the relation between the sample size and the value of the BF.

@@

As shown in Figure \@ref(fig:simulated-power), the natural logarithm of the BF in favour of the alternative hypothesis is growing proportionally with sample size. More precisely, whereas BFs computed on small samples (i.e., below 80 participants) support the null hypothesis, BFs computed on larger samples support the alternative hypothesis for all three facial muscles. For instance, the average $\text{BF}_{10}$ computed for the OOI muscle with a sample size of 160 participants is of $\exp(2.18) \approx 8.85$, indicating that these data are approximately $8.85$ times more likely under the alternative hypothesis than under the null hypothesis. To sum up, this reveals that although at low sample sizes, the BF may provide (weak) evidence for the null hypothesis (relative to the alternative hypothesis), this pattern may very well reverse for higher sample sizes.

@@

```{r order, fig.cap = "Average natural logarithm of the EMG peak amplitude by muscle, condition, and group. The green dots and intervals represent the by-group average and 95\\% confidence interval for the participants that first went through the rumination condition, then through the distraction condition. The orange dots and intervals represent the by-group average and 95\\% confidence interval for the participants that first went through the distraction condition, then through the rumination condition. The light green and orange dots in the background represent the individual-level average natural logarithm of the EMG amplitude by muscle, condition, and group."}
# position for groups
pd <- position_dodge(0.8)

@@

Unfortunately, the problems we discussed in the next section about the interpretation of under-powered non-significant results also apply to this test. Namely, obtaining a non-significant effect of group is very weak evidence that order did not play a role in the results, given the low power of the tests that were performed. This statistical argument is supported by the brief visual exploration fo the data that we presented in Figure \@ref(fig:order), which suggests possibly crucial differences between the two groups of participants. However, given the sample size in each group (N = 12 and N = 14), it is impossible to know for sure at this point.

@@

With this paper we aimed to nuance the strong conclusion made by @moffatt_inner_2020, who asserted that the inner experience of rumination was not related to its peripheral muscular correlates. First, we reanalysed the data from @moffatt_inner_2020 and provided some nuance to the conclusion that can be made from these data. Second, we discussed the statistical and epistemological reasons that cast doubt upon the main conclusion of @moffatt_inner_2020. Because the tests conducted by @moffatt_inner_2020 were heavily under-powered, they provide only weak evidence for the absence of difference. Third, we highlighted that the order of the conditions participants went through may impact the effects of the rumination induction (although we can not decide on this issue with the present data). Finally, we showed that the group analyses masked important inter-individual variability that should be more carefully examined.

@@

In @nalborczyk_dissociating_2020, we suggested that this observation was coherent with the mental-habit view of depressive rumination [@watkins_habit-goal_2014], which defines rumination as a habitual behaviour, automatically triggered by contextual cues such as negative mood. We know habitual behaviours are more automatic (they are not intentionally initiated) than non-habitual behaviours. Interestingly, it has been observed that the automaticity with which a verbal thought is evoked may influence the degree to which it is enacted, that is, the degree to which it recruits the speech motor system [e.g., @cohen_motor_1986; @sokolov_inner_1972]. According to @cohen_motor_1986, the presence of peripheral motor activity during inner speech production may be interpreted in terms of attention sharing. For instance, in novel (hence non-automatic) or difficult situations, the vividness of inner speech may be strengthened by increasing the speech motor activity, resulting in more salient auditory percepts. Relating this idea to the motor control framework we previously proposed [e.g., @loevenbruck_cognitive_2018; @grandchamp_condialint_2019], it may be said that the characteristics of the situation (e.g., novelty, difficulty) may influence the amount of inhibition that is applied to motor commands during inner speech production, hence resulting in more or less visible peripheral muscular activity [see also a discussion of theses ideas in the broader context of motor imagery, @guillot_imagining_2012].

Another possible interpretation is that automatic forms of inner speech may rely more heavily on higher-level (e.g., memory-based) cognitive processes whereas less automatic (i.e., more intentional or deliberate) forms of inner speech may rely more on simulation mechanisms via the use of internal models of the speech motor systems [@nalborczyk_understanding_2019]. In other words, the production of automatic or non-automatic inner speech would be underpinned by different processes that would involve the speech motor system to a different extent. This distinction is similar to the distinction between the two routes of predictions-by-association and prediction-by-simulations in speech perception and comprehension [@pickering_integrated_2013]. The prediction-by-association mechanism would rely more on perceptual sensory experiences and domain-general cognitive abilities whereas the prediction-by-simulation mechanism would rely more simulation of the motor action leading to the speech auditory percept. In the former case, no peripheral muscular activity is expected, whereas in the latter case, the speech motor system would be involved in simulating/emulating the corresponding overt action [cf. also the motor simulation vs. direct simulation (memory retrieval) distinction in @tian_mental_2012]. Whether the physiological correlates of automatic and non-automatic (deliberate) forms of inner speech differ because of inhibitory constraints or because they rely on different processes (e.g., prediction-by-association or prediction-by-simulation) remains an open empirical question. We previously discussed these issues in more length and suggested ways forward from an experimental perspective [cf. the discussion in @nalborczyk_understanding_2019].
